Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-25 paper.Rmd
submission revision (sent)

output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
header-includes:
  - \usepackage{floatrow}
  - \floatsetup[figure]{capposition=bottom}
  - \floatsetup[table]{capposition=top}
  - \usepackage[format=plain,labelfont=bf,up]{caption}
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

B&S made forecasted the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assess the accuracy of these forecasts. The forecasts themselves are probabilistic, i.e. range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts ("1") and the remaining forecasts to have been for no onset ("0"). They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled "Assuming Persistence"), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets ("Assuming Change").\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices. There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset. In fact, this a conscious choice B&S seem to acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.

@@

It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. Theoretically, a conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_. Rather than not making any "assumptions about the persistence of civil war", this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2. Since Sambanis is the author of the civil war data that is used, it would  already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing. It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.

Does it make sense to count a high probability for the risk of onset for a case with an ongoing civil war as a partial or complete "win" for the model? When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1"). In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also do try to claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases. This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g. two separatists conflicts in different portions of the country. But this is not the case; ongoing conflicts are consistently in all other analyses dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.

>>>>

output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
header-includes:
  - \usepackage{floatrow}
  - \floatsetup[figure]{capposition=bottom}
  - \floatsetup[table]{capposition=top}
  - \usepackage[format=plain,labelfont=bf,up]{caption}
  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

B&S forecast the risk of civil war onset for the first half of 2016 (2016-H1, where H2 would be the second half), and assessed the accuracy of these forecasts. The forecasts themselves are probabilistic, i.e. range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts ("1") and the remaining forecasts to have been for no onset ("0"). They evaluate two alternative codings of observed civil war outcomes, one in which 15 countries, including Colombia, had ongoing civil wars at the end of 2015 (labeled "Assuming Persistence"), and an alternative coding in which the civil war in Colombia ends but Turkey and Benin experience civil war onsets ("Assuming Change").\footnote{\label{fn:leadDV}The former coding is in the main replication data, while the latter is hand-coded in the Stata script that produces their confusion matrices. There is also some potential ambiguity about the discussion in this section because the forecasts B\&S claim are for the first half of 2016 are actually for the second half of 2016, but scored against outcomes for the first half of 2016.} However, when they score the forecasts to produce the confusion matrices they report in their Table 4, they score them using civil war incidence, not onset. In fact, this a conscious choice B&S seem to acknowledge in the text (if not the table with confusion matrices), but it's not clear to us how this is justifiable.

@@

It's not clear to us under what conditions forecasting civil war onset risk for countries with ongoing civil wars make sense. A
conflict could have ended in 2015-H2 and a new one started right away in 2016-H1, but when making the forecasts, _we would have already known that_. Rather than not making any "assumptions about the persistence of civil war", this would only make sense if we in fact assume that all ongoing civil wars ended in 2015-H2. Since Sambanis is the author of the civil war data that is used, it would  already have been known in early 2016 that these 15 civil wars had in fact not terminated, and were still ongoing. It is thus both by definition impossible for countries with ongoing civil wars to experience another onset, and at the time the forecasts were made it would have been known that all conflicts were still ongoing up through the beginning of the forecasting period.

Does it make sense to count a high probability for the risk of onset for a case with an ongoing civil war as a partial or complete "win" for the model? When creating the confusion matrices, B&S treat these ongoing civil war instances equivalent to civil war onsets, as positive events ("1"). In the text, B&S do acknowledge a distinction between predicting an actual onset versus predicting a continuation, but they also claim credit for the fact that the model places high probabilities on most of the ongoing civil war cases. This would only make sense if the civil war coding was at the conflict, not country-level, so that a country could experience more than one civil war at any given time, e.g. two separatists conflicts in different portions of the country. But this is not the case; ongoing conflicts in all other analyses  are consistently dropped from the data as if a new civil war onset for a country with an ongoing conflict is conceptually impossible.
